Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-10-31 index.qmd
section:introduction,issues,structural; process:write

<!-- 1. Where are we now with DCJ? -->
Over the past decade, numerous studies have documented the effectiveness of the *comparative judgment* (CJ) method [@Thurstone_1927] for assessing competencies and traits. These studies have evaluated CJ from three main perspectives: its ability to produce reliable and valid trait scores, its practical applicability, and its time efficiency. Research on reliability and validity shows that CJ can generate precise and consistent scores [@Pollitt_2012a; @Pollitt_2012b; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @Bouwer_et_al_2023] that accurately represent the traits being measured [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bouwer_et_al_2023]. Research on practical applicability highlights CJ's versatility across both educational and non-educational contexts, presenting it as an efficient and effective alternative for measurement and evaluation [@Pollitt_2004; @Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020]. Lastly, research on time efficiency suggests that CJ can offer at least equal, if not significant, time savings when evaluating stimuli compared to traditional marking methods [@Pollitt_2012a; @Pollitt_2012b; @Coertjens_et_al_2017; @Goossens_et_al_2018].

<!-- 2. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of studies on CJ, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed. This research primarily focuses on three: the apparent disconnect between CJ's structural and measurement model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role and impact of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.

@@

<!-- 1. The BTL model is CJ's measurement model -->
In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model [@Andrich_1978; @Bramley_2008]. This model specifies how latent variables are estimated from manifest variables [@Everitt_et_al_2010]. In CJ, multiple judges engage in several rounds of pairwise comparisons to assess the relative manifestation of a trait between two stimuli. Each comparison generates a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher degree of the trait. The BTL model then uses these observed outcomes to estimate scores that represent the latent trait of interest [@Pollitt_2012a; @Pollitt_2012b; @Whitehouse_2012; @Jones_2015; @vanDaal_et_al_2016; @Lesterhuis_2018; @Boonen_et_al_2020; @Bouwer_et_al_2023].

<!-- 2. But there is no structural model, only post estimation analyses -->
Moreover, researchers often use these BTL-generated scores or their transformations separately, conducting further analyses or hypothesis testing. The literature shows how these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? and what is the solution? -->
However, the statistical literature cautions against using estimated scores for separate analyses and tests. A key consideration is that BTL-generated scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010]. To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately. Thus, an integrated approach combining CJ's structural and measurement models can offer significant advantages.

@@

1. the BTL model is the statistical articulation of Case 5

  From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008

2. But this model hold the greater number of assumptions

  Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p. 269], its assumptions have become predominant in the literature, but Thurstone's Case 5 is the case that holds the greater amount of simplifying assumptions.

  @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three:

  (a) related to Case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weigthed 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additivive nature of the logit scale, which helps to cancel 'bias' judges effects.

<!-- This is the case of confounding, and @Andrich_1978; @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.  -->

<!-- possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @Arlett_2003, @Guthrie_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->

@@

  (b) related to Case 3, it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. Meaning there is an amount of correlation that it is not accounted for.

@@

  (c) related to case 5, discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.

3. What problems does this bring?

<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
<!--   a.2 It does not consider that stimuli may be nested in individuals -->

  This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003

  [Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]]{style="color:red;"}.

<!-- This ties up with the next subject -->

@@

@@

@@

**Authors' contributions:** *Conceptualization:* S.G., S.DM., T.vD., and J.M.R.E; *Methodology:* S.DM., T.vD., and J.M.R.E; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* S.G., S.DM., and T.vD.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review & editing:* S.G., S.DM., and T.vD.; *Visualization:* J.M.R.E.; *Supervision:* S.G. and S.DM.; *Project administration:* S.G. and S.DM.; *Funding acquisition:* S.G. and S.DM.

>>>>

<!-- []{style="color:red;"}. -->

<!-- 1. What does CJ do? -->
[In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence by conducting pairwise comparisons of stimuli [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher level of the trait. After conducting multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the outcomes and estimate scores that reflect the underlying trait of interest. CJ has been successfully employed in assessing the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].]{style="color:red;"}

<!-- Specifically, CJ has been used to evaluate the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023] and to assess the quality of speech [@Boonen_et_al_2020]. Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_2015], engineering design skills [@Bartholomew_et_al_2018], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowdlege [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020]. -->

<!-- 2. Where are we now with DCJ? -->
Numerous studies have documented the effectiveness of CJ in assessing various traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits being measured [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

<!-- rubrics [@Coertjens_et_al_2017; @Goossens_et_al_2018]  -->
<!-- holistic benchmark ratings [@Bouwer_et_al_2023] -->

<!-- Adaptive comparison algorithms dynamically present stimuli to judges, based on the results of previous comparisons. They aim to enhance the informativeness of comparisons while optimizing judges' time [@Bramley_2015].  -->

<!-- 3. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of studies on CJ, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed. This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.

@@

<!-- 1. The BTL model is CJ's measurement model -->
In a typical CJ study, the BTL model serves as the measurement model for CJ [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, the BTL model uses the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

<!-- 2. But there is no structural model, only post estimation analyses -->
Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis testing. The literature indicates that these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the latent trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? -->
However, the statistical literature cautions against using estimated scores to conduct additional analyses or tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty when conducting additional analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].

<!-- 4. So, what is the solution?  -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately. Thus, an integrated approach combining CJ's structural and measurement models can offer significant advantages.

@@

<!-- 1. the BTL model is the statistical articulation of Case 5 -->

<!--   From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008.  -->

<!--   What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution -->

<!--   Table with assumptions -->

<!-- 2. But this model hold the greater number of assumptions -->

<!--   Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p. 269], its assumptions have become predominant in the literature, but Thurstone's Case 5 is the case that holds the greater amount of simplifying assumptions. -->

<!--   @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three: -->

<!--   (a) related to Case 2 and case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additive nature of the logit scale, which helps to cancel 'bias' judges effects.  -->

<!--   Particularly on how CJ is used now, use the text example -->

<!--   What we are saying is that the judges' effects depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder. This, in the best scenario, cause less precise estimates, and in the worst scenario, biased estimates. possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->

@@

<!--   Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.  -->

<!--   use the example of writing!!! -->

<!--   (b) related to Case 3, it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. Meaning there is an amount of correlation that it is not accounted for.  -->

<!--   Here we loose precision, because we are not controlling for clustering -->

@@

<!--   (c) related to case 5, discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals. Example of exaggerated comparison of individuals, university students vs secondary school students. -->

<!--   here we face misspecification in the model -->

<!-- If your scores are not reliable they cannot be valid. People have been employing a model with assumptions that they do not know if those assumptions apply for their cases. And when you do that, the scores you produce most likely will not be reliable, which, in turn, renders them invalid. -->

<!--   [Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]]{style="color:red;"}. -->

@@

<!-- @Bramley_2015 -->
<!-- The implication is that the SSR statistic is worthless as an indicator of the quality / consistency / reliability of a scale constructed from an ACJ study that has involved a significant number of adaptive rounds. With random or fixed rounds, the SSR values were all below 0.25, which would generally be taken as an indicator of failure to create a meaningful scale. To put it another way, a low value of SSR almost certainly indicates low reliability, but a high value of SSR does not necessarily indicate high reliability.  -->
<!-- This is not to say that the ACJ studies listed in Table 1 (or indeed other studies not listed) were themselves worthless, just that not much should be made of claims about very high reliability in ACJ studies on the basis of high values of the SSR statistic. In fact, many of those studies also showed moderate to high correlations with variables external to the analysis, which were rightly interpreted as evidence of concurrent validity. -->
<!-- The conclusion is therefore that the SSR statistic is at best misleading and at worst worthless as an indicator of scale reliability whenever a CJ study has involved a significant amount of adaptivity. -->

@@

<!-- Due to efficiency of scoring, most of research only colects only one text. This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability. However, this cannot be known ad-hoc.  -->

@@

**AI-assisted technologies in the writing process:** The authors used ChatGPT, an AI language model, during the preparation of this work. They occasionally employed the tool to refine phrasing and optimize wording, ensuring appropriate language use and enhancing the manuscript's clarity and coherence. The authors take full responsibility for the final content of the publication.

**CRediT authorship contribution statement:** *Conceptualization:* S.G., S.DM., T.vD., and J.M.R.E; *Methodology:* S.DM., T.vD., and J.M.R.E; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* S.G., S.DM., and T.vD.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review & editing:* S.G., S.DM., and T.vD.; *Visualization:* J.M.R.E.; *Supervision:* S.G. and S.DM.; *Project administration:* S.G. and S.DM.; *Funding acquisition:* S.G. and S.DM.
